---
title: "hw03.Rmd"
author: "Gurjot Singh"
date: "01/10/2018"
output: github_document
---

# Homework 3 - singhg53

Hello and welcome to my Rmd file for homework #3. Today we will be exploring the `tidyverse` package, specifically `ggplot2` and `dplyr` packages using the `gapminder` dataset.

First lets load the dataset, `gapminder` and the `tidyverse` package which has `dplyr` the data manipulation tool and `ggplot2` the data visualization tool.


```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
```
Now let's analyze some data.

# Task 1
## Get the maximum and minimum of GDP per capita for all continents.

First lets assign the minimum and maximum GDP per capita, grouped by continents, to variable `GDPcont`.

```{r}
GDPcont <- gapminder %>%
  group_by(continent) %>%
  summarize(min = min(gdpPercap), max = max(gdpPercap))

knitr::kable(GDPcont)
```

Now let's create some plots. For the first plot we will look at the minimum GDP per capita for each continent.

```{r}
ggplot(GDPcont, aes(continent, min)) + 
  geom_bar(stat = "identity", color = "red", fill = "red") + 
  scale_y_log10() +
  ggtitle("Minimum GDP per Capita for Each Continent") + 
  ylab("Minimum GDP per Capita") +
  xlab("Continent")
```

We can do the same for the maximum GDP per capita values for each continent

```{r}
ggplot(GDPcont, aes(continent, max)) + 
  geom_bar(stat = "identity", color = "green", fill = "green") + 
  ggtitle("Maximum GDP per Capita for Each Continent") + 
  ylab("Maximum GDP per Capita") +
  xlab("Continent")
```



QUESTION: HOW TO PLOT THEM TOGETHER

# Task 2
## Look at the spread of GDP per capita within the continents.

```{r}
ggplot(gapminder, aes(continent, gdpPercap, fill=continent)) + 
  scale_y_log10() +
  geom_violin() +
  geom_boxplot() +
  geom_jitter(alpha=0.1) +
  labs(title="Violin Plot of GDP per Capita for Each Continent") +
  ylab("GDP per Capita") +
  xlab("Continent")
```

# Task 3
## Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population.

```{r}
trimmed_mean <- gapminder %>% group_by(year) %>% summarize("Untrimmed Mean LifeExp" = mean(lifeExp), "Trimmed (=0.125) Mean LifEexp" = mean(lifeExp,trim = 0.125)) %>% arrange(year)

trimmed_mean
```

trimming by 12.5% would discard the minimum and maximum value in the sample

```{r}

```

